Introduction

### 2020-08-20 code/table1.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

<div style="text-align: right"> Table 1 Data </div>
```{r table-1-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
load("../data/moderate_enrich_prob.Rda") ## load data
load("../data/gene2_level2_lookup.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")
load("../data/pathname2level2.Rda")

@@

cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% kable(caption = "Table 1: Most mutated gene in class-specific pathways (enrichment score >= 0.30) for specific cancer", row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```
<p style="FONT-SIZE:10px; COLOR:#000000; LINE-HEIGHT:12px;">**Table 1:** The numbers in this table are completely wrong, it is just a place holder, please ignore</p>

>>>>

```{r table-1-data-generation, eval = FALSE}
tumor_clusters <- fread("../data/umap_3d_coors.tsv") ## load in data
load("../data/moderate_enrich_prob.Rda") ## load data
load("../data/gene2_level2_lookup.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")
load("../data/pathname2level2.Rda")

@@

cluster_spec_enrich_table %>% .[, c("Pathway", "Class", n_tumor)] %>% knitr::kable(row.names = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 12) %>%
  column_spec(1, bold = T)
```

### 2020-08-20 text/discussion.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. And that is not accounting for the fact that
~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway as the unit of
disruption instead of the gene, employing the conceit that individual mutations
may be sufficient to disrupt or misdirect pathway activity (as is often the case
with driver mutations). Rather than evaluating the burden of mutations on each
pathway within or across tumors, which can be biased by the number of genes or
the coverage of pathway-linked genes in the genome, we scored each pathway in
binary fashion, allowing one mutation or multiple mutations to count equally for
each sample.

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-gain, untranslated region and splice site mutations in genes
that are actively expressed in each cancer type, thus avoiding bias from
transcription coupled repair. Our method of filtration differs from the
"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
account for the same biases. We also filtered genes on the basis of pathway
membership in biochemical pathways, excluding curated gene sets related to
diseases, syndromes, or classes of proteins with shared catalytic activity or
conserved domains ( _e.g._ transcription factors) which are potentially
problematic [@pathways-Atul-Butte]. We chose this approach to limit redundancy,
which is already high in the remaining pathways for some driver genes, and to
exclude biologically unrelated collections of genes. We do not account for
non-coding mutations, amplification events or structural variants, though we
hope to include them in subsequent studies (see below). Our pathway-centric
analysis also allows us to circumvent the problem that some cancer-promoting
mutations likely occur at such low frequencies as to be indistinguishable from
background. Previous attempts to capture these associations using Bayesian
networks have yielded important new insights into this problem and new genes
[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
traditional analysis, supporting the case that pathway-centric approaches have
the potential to increase statistical power.

@@

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways as represented by our classes (**figures 1 & 2**).

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the classes we identified are consistent with a
signature of overall disease progression, such that successive, adjacent classes
on the UMAP projection exhibit increasing specificity. We identified no such
trend in the number of mutations (see **figure 4**), the relative staging
(**figure 3D**) or metastasis (**figure 3E**); and each class instead was
associated with unique combinations of pathways (as in **figures 5B, C**).

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS mutant
cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary figure
S1**) are unevenly distributed, though we could not identify any cancers that
were exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 3B** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to
histological and molecular expression profiles that define them and how they
respond to experimental stimuli [@breast-subtypes2]. It has been hypothesized
that differences in the molecular regulators of development in the precursor
cell types present in the breast epithelium drive the various histological
phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 3B**).
However, both subtypes also contained samples in every other class (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one class or another but are not determinative;
this view is compatible with the previously stated hypothesis but opens the way
for a more granular view of individual tumors. It would be surprising if we did
not observe bias for some cancers and subtypes amongst our classes, since some
treatment regimens have greater efficacy for patients of a given cancer or
histological subtype [@breast-subtypes2]. Nonetheless, the basis for some tumors
being treatment-refractory in spite of receiving the standard of clinical care
for diagnostic markers remains elusive. Doubtless some of this is due to bad
luck, as tumors can metastasize and remain dormant years before they are
detected at distal sites, or resistant clones may have already arisen at
undetectable levels [@hallmarks_II], but our analysis suggests the possibility
of identifying more informative molecular, histological or cellular subtypes
that could form a basis for future separation of patients into different
precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, for which we did not find many distinctive associations with
pathways (**figure 5B, C**). Class 1 had a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 3D, E**). This class likely represents a
group of tumors with aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of structural variations _vs._ other variant types [@PCAWG-MAIN] and
are heavily skewed to cluster 1 membership (**Supplementary figure S1**).
Likewise, Class 10 represents a group of hyper-mutated tumors that harbor so
many mutations that virtually no pathway is unaffected. It seems likely that a
significant fraction of the “mutant” samples for each pathway are burdened with
excess passenger mutations. This could be addressed with application of
ever-increasingly sophisticated filtering of likely passenger mutations [ _e.g._
@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
incorporate these other data into a comprehensive pathway-centric analysis as we
have done here for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 9**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class and
cancer type fit better than models in which they are independent factors.
However, considering that within each cancer type there are different clinical
standards of care, and even within classes of drugs the preferred treatment can
very between cancers, it makes sense that we observe tissue-specific class
effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
hormonally driven cancers, for example. Ovarian cancer has but one main
treatment axis, platinum, whereas breast cancer patients have a variety of
treatment regimens based on molecular subtype and other factors. It is easy to
see how standard of care for one cancer and class could produce different
outcomes than the same class of a different cancer type. Unfortunately, given
the diversity of drug classes and treatments (including non-chemical treatments
such as surgery and palliative care), we lack sufficient power to explore these
variables in the TCGA data. It is our hope that future studies will help to
distinguish between treatment-specific effects on survival given different
pathway disruption classes.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and are frequently the result of aberrations involving common driver genes such
as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness (as in **figure 6** and **tables 1** and **ST2**). Importantly, our
observations do not imply the order in which these mutations should accumulate.
This could be addressed in a future study by integrating our analysis with
evolutionary analysis of clonality, drawing inferences from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to hypothesize that these mutations promote or enable the subsequent acquisition
of class-specific defects via random mutation and natural selection, thus
producing the classes we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as APOBEC,
BRCA1 and BRCA2 produce mutational hotspots reflecting the varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed together, and supporting the view that, within the parameters of
our analysis, metastasis has class-specific requirements. Since the number of
metastatic samples is relatively low this part of our analysis is likely
underpowered and subject to expanded analysis with larger cohorts. Nonetheless,
the class-specific metastatic pathway enrichment we observe (in **table 1**)
could be a result of the pathways directly promoting metastatic phenotypes in
these classes, or because they acquire a stronger or more complete oncogenic
phenotype, as we don’t know the order in which they were acquired. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on one or more deadly phenotypes that capture most of the critical
features of end-stage cancer. One explanation for this could be that tumors must
accumulate mutations in a certain order or face elimination due to any of the
selective bottlenecks they face (senescence, the immune system, _etc._)
[@hallmarks_II]. Some clusters were characterized by cluster-specific pathways
that seem superficially unrelated; for example cluster 2 has mutations in the
proteasome subunits, cluster 3 has mutations impacting protein translation,
whereas cluster 5 has defects in RNA metabolism and clusters 7, 8 and 9 have
defects in the more traditional extra-cellular signaling pathways (as
exemplified in **table S1**). It may be that the primary effects of mutations in
protein and RNA metabolism in clusters 3 and 5 point to currently
underappreciated mechanisms impacting hallmark pathways. Clusters 3 and 5 are
also marked by a decidedly different set of pathways in metastasis. Clusters 3
and 5 notably have (in addition to signaling pathways) mutations in glucagon and
glycogen metabolism, respectively, in their metastases (**table 1**), consistent
with energy metabolism as one of the new hallmarks [@hallmarks_II], but also
distinguishing these classes from the others in the order or nature of hallmark
acquisition. Thus, even at relatively low power, our analysis of metastasis
further supports the pathway-centric approach as a fruitful way to uncover real
differences in underlying biology. As a major caveat to this, we acknowledge
that many therapies _do_ target highly-specific driver genes, markers, and
signaling pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader
context of the genetic background and pathway vulnerability of tumors containing
such markers may aid in creating smarter combination therapies. We submit that
when we discover the requirements of each cluster with respect to pathway
disruptions and metastasis we may be able to target them therapeutically and
prevent further adaptation.

>>>>

One of the biggest hurdles in cancer research is the sparsity of data; ~20,000
protein-coding genes is comparable with the number of tumor samples, even with
multiple mutations per sample. And that is not accounting for the fact that
~30,000 non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor activity [@deesMuSiC2012;
@tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @noncoding-mutations-Zhang]. Hence, we sought to
simplify the problem by employing a "knowledge-base driven analysis"
[@pathways-Atul-Butte], investigating cancer as a disease of basic cellular and
biochemical pathways, which we accomplished by translating gene-level mutations
into pathway level disruptions. Our approach differs from previously described
methods [@PATHNET] in that we chose to focus on the pathway (defined in the
methods) as the unit of disruption instead of the gene, employing the conceit
that individual mutations may be sufficient to disrupt or misdirect pathway
activity (as is often the case with driver mutations). Rather than evaluating
the burden of mutations on each pathway within or across tumors, which can be
biased by the number of genes or the coverage of pathway-linked genes in the
genome, we scored each pathway in binary fashion, allowing one mutation or
multiple mutations to count equally for each sample.

To our knowledge, this approach has not previously been attempted in the
literature in spite of its relative simplicity. We limited our analysis to
mutations with likely deleterious effects, including missense, nonsense,
frame-shift, stop-loss, untranslated region and splice site mutations in genes
that are actively expressed in each cancer type, thus avoiding bias from
transcription coupled repair. Our method of filtration differs from the
"rank-and-cut" method of [@PCAWG-MAIN] but represents a reasonable attempt to
account for the same biases. We also filtered genes on the basis of pathway
membership in biochemical pathways, excluding curated gene sets related to
diseases, syndromes, or classes of proteins with shared catalytic activity or
conserved domains ( _e.g._ transcription factors) which are potentially
problematic [@pathways-Atul-Butte]. We chose this approach to limit redundancy,
which is already high in the remaining pathways for some driver genes, and to
exclude biologically unrelated collections of genes. We do not account for
non-coding mutations, amplification events or structural variants, though we
hope to include them in subsequent studies (see below). Our pathway-centric
analysis also allows us to circumvent the problem that some cancer-promoting
mutations likely occur at such low frequencies as to be indistinguishable from
background. Previous attempts to capture these associations using Bayesian
networks have yielded important new insights into this problem and new genes
[@hornNetSigNetworkbasedDiscovery2018] that would not have been possible with
traditional analysis, supporting the case that pathway-centric approaches have
the potential to increase statistical power.

@@

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomics, proteomics, and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways as represented by our clusters (**figures 1 & 2**).

We attempted to account for our clusters in terms of more trivial explanations.
For example, it could be that the clusters we identified are consistent with a
signature of overall disease progression, such that successive, adjacent
clusters on the UMAP projection exhibit increasing specificity. We identified no
such trend in the number of mutations (see **figure 4**), the relative staging
(**figure 3D**) or metastasis (**figure 3E**); and each cluster instead was
associated with unique combinations of pathways (as in **figures 5B, C**).

We find that most tumors segregate into one of ten biologically distinct
clusters, regardless of the cancer type (**figure 2F, 3A** and **supplementary
figure S1**). Some cancer types ( _e.g._ Pancreatic:PAAD in the RAS mutant
cluster 8, and Kidney Chromophobe:KICH in cluster 1, see **supplementary figure
S1**) are unevenly distributed, though we could not identify any cancers that
were exclusive to a single tumor cluster. Only cancer types with the fewest
samples were found to be absent from one or more clusters at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 3B** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
could reflect differences in cell-of-origin, rather than fundamental differences
in proximal cause. The four major subtypes of breast cancer correspond to
histological and molecular expression profiles that define them and how they
respond to experimental stimuli [@breast-subtypes2]. It has been hypothesized
that differences in the molecular regulators of development in the precursor
cell types present in the breast epithelium drive the various histological
phenotypes [@breast-subtypes;
@breast-subtypes3]. Consistent with this view we found that breast tumor samples
of the Luminal A subtype were heavily biased toward membership in clusters 1 and
7, and basal tumor samples were biased toward cluster 6 (**figure 3B**).
However, both subtypes also contained samples in every other cluster (without
exception), and Luminal B and Her2 positive samples are distributed across
clusters. We interpret these data to mean that inherited cell-of-origin
molecular profiles may predispose certain precursor cell types within the breast
epithelium to forming tumors of one cluster or another but are not
determinative; this view is compatible with the previously stated hypothesis but
opens the way for a more granular view of individual tumors. It would be
surprising if we did not observe bias for some cancers and subtypes amongst our
classes, since some treatment regimens have greater efficacy for patients of a
given cancer or histological subtype [@breast-subtypes2]. Nonetheless, the basis
for some tumors being treatment-refractory in spite of receiving the standard of
clinical care for diagnostic markers remains elusive. Doubtless some of this is
due to bad luck, as tumors can metastasize and remain dormant years before they
are detected at distal sites, or resistant clones may have already arisen at
undetectable levels [@hallmarks_II], but our analysis suggests the possibility
of identifying more informative molecular, histological or cellular subtypes
that could form a basis for future separation of patients into different
precision treatment regimens.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
clusters, 1 and 10, for which we did not find many distinctive associations with
pathways (**figure 5B, C**). Cluster 1 had a relatively low proportion of
mutated pathways across the board, although it is broadly enriched in many of
the same tumor-promoting pathways common to the other groups. This likely does
not reflect a true difference in stages of tumor evolution, as our data show
clearly that this cluster is as likely to contain stage IV metastatic tumors as
it is to contain those of stage I (**figure 3D, E**). This cluster likely
represents a group of tumors with aberrations in methylation, copy number
alterations, structural variants, or combinations thereof which we excluded from
our analysis. Consistent with this, kidney chromophobe and thyroid cancers have
high proportions of structural variations _vs._ other variant types
[@PCAWG-MAIN] and are heavily skewed to cluster 1 membership (**Supplementary
figure S1**). Likewise, cluster 10 represents a group of hyper-mutated tumors
that harbor so many mutations that virtually no pathway is unaffected. It seems
likely that a significant fraction of the “mutant” samples for each pathway are
burdened with excess passenger mutations. This could be addressed with
application of ever-increasingly sophisticated filtering of likely passenger
mutations [ _e.g._
@spliceAI; @deep-learning-clinical-variants]. In the future, we hope to
incorporate these other data into a comprehensive pathway-centric analysis as we
have done here for point mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption clusters exhibit cancer-type specific effects on
expected survival (**figure 9**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on cluster and
cancer type fit better than models in which they are independent factors.
However, considering that within each cancer type there are different clinical
standards of care, and even within classes of drugs the preferred treatment can
very between cancers, it makes sense that we observe tissue-specific cluster
effects. Contrast the situation with ovarian _vs._ breast cancer, which are both
hormonally driven cancers, for example. Ovarian cancer has but one main
treatment axis, platinum, whereas breast cancer patients have a variety of
treatment regimens based on molecular subtype and other factors. It is easy to
see how standard of care for one cancer and cluster could produce different
outcomes than the same cluster of a different cancer type. Unfortunately, given
the diversity of drug classes and treatments (including non-chemical treatments
such as surgery and palliative care), we lack sufficient power to explore these
variables in the TCGA data. It is our hope that future studies will help to
distinguish between treatment-specific effects on survival given different
pathway disruption clusters.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, cluster-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined clusters (**supplementary figure S1**)
and are frequently the result of aberrations involving common driver genes such
as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Cluster-specific
evolution must involve the acquisition of disruptions to pathways that may
individually be harmful ( _e.g._ highly proliferative cells are more likely to
senesce) but together produce more specialized cancer phenotype and increased
fitness (as in **figure 6** and **tables 1** and **ST2**). Importantly, our
observations do not imply the order in which these mutations should accumulate.
This could be addressed in a future study by integrating our analysis with
evolutionary analysis of clonality, drawing inferences from variant allele
frequencies as in @PCAWG-EVO. However, since many of the genes in the
non-cluster-specific pathways involve the known driver genes, it is reasonable
to hypothesize that these mutations promote or enable the subsequent acquisition
of cluster-specific defects via random mutation and natural selection, thus
producing the clusters we observed. In support of this, the pan-cancer analysis
of whole genomes consortium (PCAWG) found that oncogenic driver mutations are
highly enriched in early arising clones, whereas later arising clones have much
greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes that
are known to be responsible for discrete mutation signatures such as APOBEC,
BRCA1 and BRCA2 produce mutational hotspots reflecting the varying selective
pressures in different tissues [@PCAWG-MAIN].

@@

We report that pathway enrichment in metastatic tumors across all clusters
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed together, and supporting the view that, within the parameters of
our analysis, metastasis has cluster-specific requirements. Since the number of
metastatic samples is relatively low this part of our analysis is likely
underpowered and subject to expanded analysis with larger cohorts. Nonetheless,
the cluster-specific metastatic pathway enrichment we observe (in **table 1**)
could be a result of the pathways directly promoting metastatic phenotypes in
these clusters, or because they acquire a stronger or more complete oncogenic
phenotype, as we don’t know the order in which they were acquired. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar clusters
converging on one or more deadly phenotypes that capture most of the critical
features of end-stage cancer. One explanation for this could be that tumors must
accumulate mutations in a certain order or face elimination due to any of the
selective bottlenecks they face (senescence, the immune system, _etc._)
[@hallmarks_II]. Some clusters were characterized by cluster-specific pathways
that seem superficially unrelated; for example cluster 2 has mutations in the
proteasome subunits, cluster 3 has mutations impacting protein translation,
whereas cluster 5 has defects in RNA metabolism and clusters 7, 8 and 9 have
defects in the more traditional extra-cellular signaling pathways (as
exemplified in **table S1**). It may be that the primary effects of mutations in
protein and RNA metabolism in clusters 3 and 5 point to currently
underappreciated mechanisms impacting hallmark pathways. Clusters 3 and 5 are
also marked by a decidedly different set of pathways in metastasis. Clusters 3
and 5 notably have (in addition to signaling pathways) mutations in glucagon and
glycogen metabolism, respectively, in their metastases (**table 1**), consistent
with energy metabolism as one of the new hallmarks [@hallmarks_II], but also
distinguishing these clusters from the others in the order or nature of hallmark
acquisition. Thus, even at relatively low power, our analysis of metastasis
further supports the pathway-centric approach as a fruitful way to uncover real
differences in underlying biology. As a major caveat to this, we acknowledge
that many therapies _do_ target highly-specific driver genes, markers, and
signaling pathways ( _e.g._ TP53, EGFR or HER2), but understanding the broader
context of the genetic background and pathway vulnerability of tumors containing
such markers may aid in creating smarter combination therapies. We submit that
when we discover the requirements of each cluster with respect to pathway
disruptions and metastasis we may be able to target them therapeutically and
prevent further adaptation.

### 2020-08-20 text/introduction.Rmd
respond to critiques from David Van Valen, Kate Lawrenson, Simon Knott (excluding survival section)

Recent advances in technology have made the study of cancer possible at
unprecedented levels of detail and scale. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations. To take
advantage of these tools several large consortia, including the cancer genome
atlas (TCGA) and pan-cancer analysis of whole genomes (PCAWG), sequenced large
numbers of tumors and collected data from multiple assays that could be analyzed
together, with the goal of integration and increased understanding of the
mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has
been estimated that on the order of three mutations in key oncogenes and/or
tumor suppressors would be sufficient to transform a normal cell to a cancerous
state [@vogelsteinPathCancerThree2015].

Other studies aim at providing a comprehensive view of mutations, gene
expression and genomic signatures, with the express goal of understanding the
common themes of all cancer independent of tissue of origin. Understanding
cancer as a disease of the cell has long been a goal of the field and is perhaps
best characterized in essays by @weinberg2000; @hallmarks_II (see also
Discussion). The first of these studies in the genomic era identified 11 subtypes
from 12 cancer types, using integrative analysis with co-equal weighting of gene
expression, methylation, copy number and proteomics data
[@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify
28 clusters that could be further subdivided into organ specific groups,
including pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/ squamous and pan-kidney
[@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH
carried out near-exhaustive analysis of pathway enrichment using whole genome
data. Horn _et al._ [@hornNetSigNetworkbasedDiscovery2018] introduce approaches
to address the sparseness of mutations, allowing the potential importance of a
gene to be influenced by the aggregate of mutations in its nearest network
neighbors. This method is powerful especially when applied to a small number (on
the order of hundreds) of tumors. One of the challenges in the field remains
that beyond a couple hundred common driver genes and mutations, there are likely
thousands of moderate effect genes that occur at such low frequency--given
tissue diversity--and low sample numbers that they are impossible to detect
using positive selection theory. Some groups have attempted to address this
challenge by applying machine learning to cancer data sets to discover groups of
functionally related genes as they interact with larger pathways and networks
[@kimPancancerAnalysisSomatic2018; @mourikis-ml-pathways; @moonlight]. These
studies have proven very effective at highlighting some potentially fundamental
disease phenotypes at the pathway level that cross cancers with different
origins at the cellular and tissue level.

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and equates gene-level
mutations to cell-biological pathway disruptions. We use this heuristic to
evaluate all cancers and show that, surprisingly, tumors fall into categories of
common pathway disruptions with distinctive origins arising from tissue-specific
gene alterations and having unique prognosis in each cancer type.

>>>>

Recent advances in technology have greatly expanded our view into the mechanisms
of cancer at the systems biology level. Array and next-generation sequencing
technologies have made it possible to assay germline and somatic mutations,
expression profiles, DNA methylation, and copy number variations from the same
tissue samples. To take advantage of these tools several large consortia,
including the cancer genome atlas (TCGA) and pan-cancer analysis of whole
genomes (PCAWG), sequenced large numbers of tumors and collected data from
multiple assays that could be analyzed together, with the goal of integration
and increased understanding of the mechanisms of cancer.

Considerable progress has been made analyzing these data. Statistical analyses
identify hundreds of global and tissue-specific cancer driver genes
[@deesMuSiC2012; @tamboreroOncodriveCLUST2013;
@lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @tokheimEvaluatingEvaluationCancer2016;
@jiangWITER2019; @zhaoModelingPositive2019] using approaches aimed at detecting
when genes are mutated at a greater rate than expected due to chance. It has
been estimated that fewer than five mutations in key oncogenes and/or tumor
suppressors would be sufficient to transform a normal cell to a cancerous state
[@vogelsteinPathCancerThree2015; @Iranzo-Network]. Somewhat paradoxically,
mutations in cancer driver genes are commonplace in somatic tissues and
correlate with age and environmental exposures [@mutations-healthy-tissues].

Other studies provide a comprehensive view of mutations, gene expression and
genomic signatures, with the express goal of understanding the common themes of
all cancer independent of tissue of origin. Understanding cancer as a disease of
the cell has long been a goal of the field as characterized in essays by
@weinberg2000; @hallmarks_II. The first of these studies in the genomic era
identified 11 subtypes from 12 cancer types, using integrative analysis with
co-equal weighting of gene expression, methylation, copy number and proteomics
data [@hoadleyMultiplatformAnalysis2014]. The principal finding was that
tissue-of-origin is the predominant driving factor, though a significant
proportion of tumors (~10%) could be reclassified independent of
tissue-of-origin. In a second study involving 33 cancer types and a much greater
number of tumors [@shenIntegrativeClusteringMultiple2009], the authors identify
28 clusters that could be further subdivided into organ specific groups,
including pan-gastrointestinal, pan-gynecological, pan-squamous,
pan-gynecological/squamous and pan-kidney [@hoadleyCellofOriginPatternsDominate2018].

More recently, attempts have been made to consider pan-cancer mutations and
other data as disruptions of normal pathway activity. @PCAWG-PATH carried out
near-exhaustive analysis of pathway enrichment using whole genome data. Network
approaches address the sparseness of mutations, allowing genes to be influenced
by the aggregate of mutations in the nearest network neighbors
[@hornNetSigNetworkbasedDiscovery2018; @Iranzo-Network]. These methods are
powerful especially when applied to a small number (on the order of hundreds) of
tumors. One of the challenges in the field remains that beyond a couple hundred
common driver genes and mutations, there may be thousands of moderate effect
genes that occur at such low frequency--given tissue diversity--and low sample
numbers that they are impossible to detect using positive selection theory. Some
groups have attempted to address this challenge by applying machine learning to
cancer data sets to discover groups of functionally related genes as they
interact with larger pathways and networks [@kimPancancerAnalysisSomatic2018;
@mourikis-ml-pathways; @moonlight]. These studies have proven very effective at
highlighting some potentially fundamental disease phenotypes at the pathway
level that cross cancers with different origins at the cellular and tissue
level.

In this study, we attempt to address these questions by focusing on discrete
molecular pathways as the unit of disruption in cancer. Our approach strips all
cell-type-specific information from the mutation data and equates gene-level
mutations to cell-biological pathway disruptions. We use this heuristic to
evaluate all cancers and show that, surprisingly, tumors that exhibit
tissue-specific gene mutation patterns nonetheless fall into common categories
of pathway disruption having unique prognoses in each cancer type.
